Composition-climate paper

### 2015-10-20 Composition_Ecology.Rmd
_Main `Rmd`_ Minor edits to the abstract bullet points.  Added an extent value and built some of the plots a bit better.
Added more references to the introduction, cleaned up and extend the Methods section.  Added blocked sample figure with color legend.

Some editing of results.

_bib file_
Added references here.  Fairly extensive.

*`comp_ecol.R`* is the purled version of the paper.

---
title: "Composition-climate paper"
author: "Simon Goring *et al*."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    pandoc_args: -V geometry:vmargin=1in -V geometry:hmargin=1in
  md_document:
    variant: markdown_github
  word_document:
    fig_height: 6
    fig_width: 6
csl: ecology.csl
pandoc_args: -H margins.sty
bibliography: goringetal_references.bib
---

@@

Abstract
====
* Much of our understanding of species-cliamte relationships and functional-community relationships in forests is based off of contemporary observations.
* Ecosystems are highly dynamic and contemporary forest have a heavy imprint of land use.  Climates also changing.  
* Hence the question of whether species-climate relationships today are representative compared to those in past.
* Here we assess stability of species-climate relationships. Using historic PLS data and FIA data.
* We show that species-climate relationships have significantly shifted for many taxa.
* Contemporary trees have shifted to cooler maximum temperatures, warmer winter temperatures, reduced seasonal temperature range, and wetter conditions.
* Relative importance of land use and climate change varies by taxon and climate variable.  
* Shifts in warmer temperature mainly due to land use change, etc.
* Implications sentence or two at the end.

@@

land_use_plot <- natural[[2]] + 
  geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c'))

@@

grid.arrange(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4))
```

@@

In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.  Goring et al [@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, but did not examine how these shifts might affect the relationships between climate and distribution for individual taxa. Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests, we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.  Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.

@@

Methods
===============
We pair pre-settlement vegetation data from the Public Land Survey System aggregated to a 64km^2^ grid [@goring2015preset] with annually resolved climate data for the region.

**Paragraph about data scale between PLS and FIA**

**Paragraph about FIA/PLS absence not really Absence**

The analysis follows a blocked sampling method.  We generate probability density functions for actual PLSS presence, by taxon, at the settlement era and for the same PLS taxa (and point locations) in the modern era.  We also generate climate data for the FIA presence, and we generate climate space for modern land use.  In this way we can test how much of the change in shape is the result of changing land use vs changing climate since the early part of the century.

We also develop estimates of the modern climate space that taxa would occupy if they remained in place, with only climate changing, and vice versa (need better explanation here).

@@

The choice of a normal for the domain is potentially problematic.  The time transgressive nature of the PLS Survey means that some of the PLS data comes from records sampled in the early 1800s, while other come from the early 1900s. Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period. As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice [ADD JULY temperatures here]. **Talk about modern climate normal & historical climate normal make it clear that this is the best we can do, and that it's good.**

@@

```{r loss_gain_fig, warning=FALSE, message=FALSE, echo=FALSE}

@@

@@

```{r densities, warning=FALSE, message=FALSE, echo=FALSE}

@@

```{r climate_shifts, warning=FALSE, message=FALSE, echo=FALSE}

>>>>

---
title: "Composition-climate paper"
author: "Simon Goring *et al*."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document:
    fig_height: 6
    fig_width: 6
  pdf_document:
    pandoc_args: -V geometry:vmargin=1in -V geometry:hmargin=1in
  md_document:
    variant: markdown_github
csl: ecology.csl
pandoc_args: -H margins.sty
bibliography: goringetal_references.bib
---

@@

Abstract
====
* Much of our understanding of species-climate relationships and functional-community relationships in forests is based on contemporary observations of tree distributions.
* Ecosystems are highly dynamic and contemporary forest have a heavy imprint of land use, while climate changes over decadal and centennial scales.
* Given long time-scale climate change and continued land use conversion, are species-climate relationships today representative of those in past.
* We assess the stability of species-climate relationships using historic PLS data and FIA data.
* We show that species-climate relationships have significantly shifted for many taxa.
* Contemporary trees have shifted to cooler maximum temperatures, warmer winter temperatures, reduced seasonal temperature range, and wetter conditions.
* Relative importance of land use and climate change varies by taxon and climate variable.  
* Shifts in warmer temperature mainly due to land use change, etc.
* Implications sentence or two at the end.

@@

ext <- c(-100000, 1050000, 600000, 1600000)

land_use_plot <- natural[[2]] + 
  geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
  coord_equal(xlim = ext[1:2], ylim=ext[3:4]) + 
  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c')) +
  theme(legend.position = 'none')

@@

grid.arrange(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4))

@@

In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.  Goring et al. [@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies,@schulte2001original,@schulte2007homogenization,@radeloff2000historical]. To date researchers have not examined how these shifts might affect the relationships between climate and the distributions of individual taxa. Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies], we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.  Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.

@@

Methods
===============
We pair pre-settlement vegetation data from the Public Land Survey System (PLSS) aggregated to a 64 $km^{2}$ grid [@goring2015preset] with annually resolved climate data for the region.  The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.  At each survey point surveyors noted the closest two to four trees, recorded their distances to the plot center point, diameters and the azimuth to the tree.  Goring et al. [-@goring2015preset] aggregated this data to an 8 x 8 km grid to reduce local-scale variability, resulting in a gridded data product with an average of 61 plots per cell, or approximately 120 trees per cell.  When examining relationships between FIA and PLS data, Goring et al. [-@goring2015] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.

Another complication is the structure of the FIA dataset.  While the PLS data was collected across the western US on a uniform grid the FIA data is only collected from forested or partially forested landscapes [@woudenberg2010forest].  As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA. This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree,@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations. [needs refs].

The analysis presented here follows a blocked sampling method (Figure 2). We generate probability density functions along climate gradients for actual PLSS presence, by taxon, at the settlement era and for the same PLS taxa (and point locations) in the modern era. We do the same for FIA presence, and for anthropogenic land use classes. Climate data is derived from monthly PRISM data from 1895 - 2014. A 30-year climate was generated for modern July temperature, January temperature, the continentality (difference between T_jul_ and T_jan_), and annual precipitation.

![model grid](figures/model_grid.png)

**Figure 2**. *Blocked method for analysis.  Data presented represent one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).  A diagonal shift from the top left to bottom right represents the true change in forest-climate relationships.*

The choice of a thirty year normal for the PLS data is potentially problematic.  The time transgressive nature of the PLS Survey means that some of the PLS data was sampled in the early 1800s, while other come from the early 1900s [@goring2015preset, @white1983history], and for much of this time period no climatological records exist. Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period. As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice.  Regional-scale reconstructions from tree-rings and from pollen-based climate reconstructions [@trouet, @mann] tell us _THIS_.

By generating kernel density estimates of taxon presence along climate gradients drawn from each of the four states of climate and vegetation (Figure 2) we can use Hellinger distance to estimate the effects of changing climate (d_H_ between kernel densities in the same column of Figure 2), or the effect of changing land use (d_H_ between kernel densities in the same row of Figure 2).  In particular, the effect of changing land use and climate (d_H_ between the top left and bottom right in Figure 2) can be parsed as the sum of differences in the left hand column (testing climate change) and the top row (testing land use only).  The relative magnitudes of these differences gives us an indication of the influence of either land use or climate in driving shifts in species distributions along the climate gradients in the Upper Midwest.

@@

@@

```{r loss_gain_fig, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}

@@

loss_table[,2:3] <- paste0(round(loss_table[,2:3]*100,1),'%')

@@

```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}

@@

```{r climate_shifts, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
